\bf ISODS: An ISODATA-Based Initial Centroid Algorithm

### 2021-01-08 paper/paper.tex
missed title update
experiment 1 table and results

\title{\bf Patent Duration, Innovative Performance, and Technology Diffusion}
\author{Noah Gardner$^1$,
      Sahar Yarmohammadtoosky$^2$, \\
      Anthony Phan$^3$,
      and Chih-Cheng Hung$^4$\\[1ex]
      \sl \small Kennesaw State University,
      Marietta, GA, USA$^{1, 3, 4}$\\
      \sl \small Iran University of Science and Technology,
      Tehran, Iran$^2$\\[1ex]
      \small $^1$ngardn10@students.kennesaw.edu, 
      \small $^2$yarmohamadishr@gmail.com, \\
      \small $^3$aphan5@students.kennesaw.edu,
      \small $^4$chung1@kennesaw.edu, \\}
\date{}
\maketitle

\begin{abstract}
      \noindent 
      Iterative Self-Organizing Data Analysis Technique (ISODATA) is a clustering
      algorithm based on K-Means that intends to find the correct number of
      clusters through merging and splitting. Compared to K-Means, the ISODATA
      algorithm should be resilient to initial cluster centers. However, ISODATA
      is also sensitive to initial cluster centers which impacts the performance
      of the algorithm. This work proposes using the ISODATA approach of splitting
      clusters to select quality initial cluster centers. The proposed approach is
      evaluated on several benchmark UCI datasets as well as the Salinas-A
      hyperspectral image dataset. The results are evaluated using standard
      clustering evaluation parameters. \\
      {\bf Keywords :} Clustering, ISODATA, Initial Centroid, Remote Sensing
\end{abstract}

@@

There are some disadvantages to the ISODATA algorithm. First, ISODATA is
sensitive to the parameters $\theta n$, $\theta e$, and $\theta c$. These
parameters should be fine-tuned for optimal performance based on the attributes
of the dataset being tested. In addition, similar to K-Means, it is sensitive to
initial cluster centers, although less so than K-Means due to the extra
processes of discarding, splitting, and merging. The initial centroids can
affect both the quality of clusters produced and the number of iterations
required to converge [CITE].

@@

The proposed clustering algorithm's objective is to improve the result we obtain
from the ISODATA algorithm by targeting the initial cluster centers. Inspired by
the ISODATA algorithm, Iterative Self-Organizing Data Split (ISODS) uses the
splitting process from ISODATA to create the required initial cluster centers to
use for the ISODATA algorithm. The inputs for the algorithm are the number of
cluster centers $K$ as an integer, and the data to generate centers from. The
algorithm starts by selecting $1$ sample as an initial cluster center. In the
next step, each data is assigned to the cluster and the mean of the cluster is
updated. Then, by the ISODATA splitting process, each cluster is split along the
dimension with the highest standard deviation. The assignments for the new
clusters are calculated based on distance and the centers are updated. This
process is continued until either the number of clusters is greater than or
equal to $K$. If the number of clusters is greater than $K$, the clusters with
the fewest members are discarded one at a time, and the assignments and centers
are updated after each rejection. The algorithm is outlined below:

@@

@@

For experiments 1 and 2, we explore applying ISODS to ISODATA and K-Means by
using benchmark UCI datasets Iris, Wine, and Glass. We also show results from
applying Random Centers and K-Means++ Centers. To evaluate the cluster results,
external cluster validation measures are used. We utilize pair-counting measures
which rely on counting matching pairs between two sets of data to determine
agreement and disagreement \cite{Rezaei2016}. The pair-counting measures applied
are \textit{rand index} (RI - agreement probability), \textit{adjusted rand
      index} (ARI), \textit{mirkin's index} (MI - disagreement probability), and
\textit{hubert's index} (HI - $P(agree) - P(disagree)$)\cite{Hubert1985,
      Meila2003}.

@@

\subsection*{Acknowledgements}
Put Acknowledgements here.

>>>>

\title{\bf ISODS: An ISODATA-Based Initial Centroid Algorithm}
\author{Noah Gardner$^1$, Sahar Yarmohammadtoosky$^2$, \\
      Anthony Phan$^3$, and Chih-Cheng Hung$^4$\\[1ex]
      \sl \small Kennesaw State University, Marietta, GA, USA$^{1, 3, 4}$\\
      \sl \small Iran University of Science and Technology, Tehran,
      Iran$^2$\\[1ex]
      \small $^1$ngardn10@students.kennesaw.edu, \small
      $^2$yarmohamadishr@gmail.com, \\
      \small $^3$aphan5@students.kennesaw.edu, \small $^4$chung1@kennesaw.edu,
      \\}
\date{}
\maketitle

\begin{abstract}
      \noindent 
      Iterative Self-Organizing Data Analysis Technique (ISODATA) is a
      clustering algorithm based on K-Means that intends to find the correct
      number of clusters through merging and splitting. Compared to K-Means, the
      ISODATA algorithm should be resilient to initial cluster centers. However,
      ISODATA is also sensitive to initial cluster centers which impacts the
      performance of the algorithm. This work proposes using the ISODATA
      approach of splitting clusters to select quality initial cluster centers.
      The proposed approach is evaluated on several benchmark UCI datasets as
      well as the Salinas-A hyperspectral image dataset. The results are
      evaluated using standard clustering evaluation parameters. \\
      {\bf Keywords :} Clustering, ISODATA, Initial Centroids, Remote Sensing
\end{abstract}

@@

There are some disadvantages to the ISODATA algorithm. First, it is sensitive to
the parameters $\theta n$, $\theta e$, and $\theta c$. These parameters should
be fine-tuned for optimal performance based on the attributes of the dataset
being tested. In addition, similar to K-Means, it is sensitive to initial
cluster centers, although less so than K-Means due to the extra processes of
discarding, splitting, and merging. The initial centroids can affect both the
quality of clusters produced and the number of iterations required to converge
      [CITE].

@@

The proposed clustering algorithm's objective is to improve the result we obtain
from the clustering algorithms by targeting the initial cluster centers.
Inspired by the ISODATA algorithm, Iterative Self-Organizing Data Split (ISODS)
uses the splitting process from ISODATA to create the required initial cluster
centers to use for a clustering algorithm. The inputs for the algorithm are the
number of cluster centers $K$ as an integer, and the data to generate centers
from.

The algorithm starts by selecting $1$ sample as an initial cluster center. In
the next step, each data is assigned to the cluster and the mean of the cluster
is updated. Then, by the ISODATA splitting process, each cluster is split along
the dimension with the highest standard deviation. The assignments for the new
clusters are calculated based on distance and the centers are updated. This
process is continued until either the number of clusters is greater than or
equal to $K$. If the number of clusters is greater than $K$, the clusters with
the fewest members are discarded one at a time, and the assignments and centers
are updated after each rejection. The algorithm is outlined below:

@@

\begin{table}
      \begin{tabular}{ |p{1.25cm}||p{1.25cm} p{1.25cm} p{1.25cm} p{1.25cm} p{1.25cm} p{1.25cm}|}
            \hline
            \multicolumn{7}{|c|}{\textbf{ISODATA}}                               \\
            \hline
            Init. & Iter.  & Clusters & ARI             & RI    & MI    & HI     \\
            \hline
            \multicolumn{7}{|c|}{\textit{Iris}}                                  \\[1ex]
            ISODS & 13.46  & 2.599    & \textbf{0.646}  & 0.826 & 0.173 & 0.653  \\
            KM++  & 15.43  & 2.544    & 0.628           & 0.816 & 0.184 & 0.631  \\
            Rand  & 11.90  & 2.41     & 0.607           & 0.802 & 0.198 & 0.604  \\
            \hline
            \multicolumn{7}{|c|}{\textit{Glass (Normalized)}}                    \\[1ex]
            ISODS & 3.19   & 1.066    & \textbf{0.0018} & 0.266 & 0.734 & -0.468 \\
            KM++  & 4.02   & 1.028    & 0.0011          & 0.265 & 0.735 & -0.470 \\
            Rand  & 3.69   & 1.055    & 0.0013          & 0.266 & 0.733 & -0.467 \\
            \hline
            \multicolumn{7}{|c|}{\textit{Wine}}                                  \\[1ex]
            ISODS & 9.16   & 1.836    & \textbf{0.313}  & 0.618 & 0.382 & 0.236  \\
            KM++  & 8.76   & 1.816    & 0.307           & 0.612 & 0.388 & 0.224  \\
            Rand  & 9.26   & 1.810    & 0.304           & 0.608 & 0.391 & 0.217  \\
            \hline 
            \multicolumn{7}{|c|}{\textbf{K-Means}}                               \\
            \hline
            \multicolumn{7}{|c|}{\textit{Iris}}                                  \\[1ex]
            ISODS & 2.29   & 3        & \textbf{0.623}  & 0.833 & 0.167 & 0.667  \\
            KM++  & 4.149  & 3        & 0.584           & 0.811 & 0.189 & 0.621  \\
            Rand  & 29.402 & 3        & 0.569           & 0.797 & 0.203 & 0.593  \\
            \hline
            \multicolumn{7}{|c|}{\textit{Glass (Normalized)}}                    \\[1ex]
            ISODS & 5.651  & 6        & 0.164           & 0.665 & 0.335 & 0.329  \\
            KM++  & 4.866  & 6        & \textbf{0.173}  & 0.643 & 0.357 & 0.285  \\
            Rand  & 77.178 & 6        & 0.151           & 0.634 & 0.366 & 0.268  \\
            \hline
            \multicolumn{7}{|c|}{\textit{Wine}}                                  \\[1ex]
            ISODS & 3.145  & 3        & \textbf{0.882}  & 0.947 & 0.053 & 0.895  \\
            KM++  & 5.388  & 3        & 0.839           & 0.926 & 0.074 & 0.851  \\
            Rand  & 24.775 & 3        & 0.834           & 0.920 & 0.080 & 0.840  \\
            \hline
      \end{tabular}
      \caption{UCI benchmark dataset results averaged over 1000 iterations of
            ISODATA and K-Means comparing three different initial cluster center
            algorithms (\textit{Init.}). Results in the table contain
            information on average iterations required (\textit{Iter.}), average
            number of clusters created (\textit{Clusters}), and validity
            measures.}
      \label{tab:ucitable}
\end{table}

@@

For the first experimental results shown in Table \ref{tab:ucitable}, we explore
applying ISODS to ISODATA and K-Means by using benchmark UCI \textit{Iris,}
\textit{Wine,} and \textit{Glass} datasets \cite{Dua2019}. We also show results
from applying Random Centers and K-Means++ Centers. For the \textit{Glass}
dataset, the data was Z-Score normalized (based on Z-Score of the training data)
because ISODATA can have problems converging on the raw dataset. We used $80\%$
of the data for training and $20\%$ of the data for testing to evaluate the
models. Following standard practice, the models were trained on the training
data, then fit to the test data. For the ISODATA parameters, we used the
approach of summarizing the class data based on the information in the training
data. The output labels (cluster assignments) were compared to the ground truth
labels contained in the test data using pair-counting measures.

To evaluate the cluster results, external cluster validation measures are used.
We utilize pair-counting measures which rely on counting matching pairs between
two sets of data to determine agreement and disagreement \cite{Rezaei2016}. The
pair-counting measures applied are \textit{rand index} (RI - agreement
probability), \textit{adjusted rand index} (ARI), \textit{mirkin's index} (MI -
disagreement probability), and \textit{hubert's index} (HI - $(P(agree)$ -
$P(disagree))$) \cite{Hubert1985, Meila2003}. When comparing ground truth labels
to cluster assignments, RI can serve as an estimation of the accuracy. ARI is
similar to the RI, though it is corrected for chance and thus does not count
randomly generated labels \cite{Rand1971}.

@@

\subsection*{Acknowledgements}
This research was supported by the College of Computing and Software Engineering
at Kennesaw State University. We also want to thank our Laboratory for Machine
Vision and Security Research for providing us resources and mentoring.
